A comparison of single-cell trajectory inference methods: towards more accurate and robust tools

### 2018-04-05 analysis/paper/paper.Rmd
update paper

<!-- A common trajectory analysis framework -->
Critical to the broad application of TI methods is that both the user's input and the method's output is standardised, so that users can easily and quickly try out different methods, compare their output, and run common analysis pipelines such as network inference [@aibar_scenicsinglecellregulatory_2017a] or module detection [@saelens_comprehensiveevaluationmodule_2018]. A common framework to analyse trajectories is especially important when the output of methods contradict, as they often do. We illustrate this in  `r ref("fig", "example_predictions")`. Here, we predicted trajectories on a datasets containing fibroblasts which were reprogrammed towards neurons and myocytes, a bifurcating trajectory [@treutlein_dissectingdirectreprogramming_2016]. The topology of all five methods differs greatly, with Slingshot underestimating the complexity of the topology, while AGA pseudotime and Monocle DDRTree overestimated the complexity, at least compared to what was expected in the data.

```{r echo=FALSE, results="asis"}
add_fig(
  fig_path = figure_file("bifurcating_example.svg", experiment_id = "14-example_predictions"),
  ref_id = "example_predictions", 
  caption_main = "An example gold standard and prediction.", 
  caption_text = "As an illustration to our practical guidelines"
)
```

@@

```{r}
add_sfig(
  figure_file("edge_flip_comparison.svg", experiment_id="3-metric_characterisation/4-edge_flip"), 
  "edge_flip_comparison", 
  paste0("Edge flip score values for common trajectory topologies"), 
  15, 
  5
)
```

>>>>

<!-- A common trajectory analysis framework -->
Critical to the broad applicability of TI methods is standardisation of the input and output interfaces of TI methods, so that users can effortlessly execute TI methods on their dataset of interest, compare different predicted trajectories, and apply downstream analyses such as network inference [@aibar_scenicsinglecellregulatory_2017a] or module detection [@saelens_comprehensiveevaluationmodule_2018]. Our framework is an initial attempt at tackling this problem, and we illustrate its usefulness here by comparing the predicted trajectories of several top-performing methods on a randomly picked dataset (`r ref("fig", "example_predictions")`). In the future, this framework should be extended to allow easier downstream analyses and also to allow easier interpretation by overlaying custom values such as cell type or expression values. In addition, further debate is required in order to arrive at a consensus concerning a common interface for trajectory models.

```{r echo=FALSE, results="asis"}
add_fig(
  fig_path = figure_file("bifurcating_example.svg", experiment_id = "14-example_predictions"),
  ref_id = "example_predictions", 
  caption_main = "Demonstration of how a common framework for TI methods facilitates broad applicability using an example dataset.", 
  caption_text = "a) The dataset consists of fibroblasts being reprogrammed towards neurons and myocytes, and thus contains a bifurcating dynamic process  [@treutlein_dissectingdirectreprogramming_2016]. b) A common visualisation of the predicted trajectories allows to easily compare the similarities and differences between predictions. The predictions are in line with earlier results in that Slingshot underestimates the complexity of the topology, AGA Pseudotime and Monocle DDRTree overestimate the complexity, and TSCAN is able to correctly identify the trajectory topology but obtains a suboptimal ordering of the cells. c) More insight can be gained into which cells ended up on which branch by placing the trajectory segments one after the other. d) Method-specific visualisations provide insight into the inner workings of a method's methodology and into how the trajectory was constructed. e) While only TSCAN and cellTree correctly predict the topology, the ordering of Slingshot and AGA Pseudotime on this dataset corresponds better to the expected ordering."
)
```

@@

```{r}
add_sfig(
  figure_file("edge_flip_comparison.svg", experiment_id="3-metric_characterisation/4-edge_flip"), 
  "edge_flip_comparison", 
 "Edge flip score values for common trajectory topologies",
  ""
)
```
